---
title: "Leaf Angle"
author: "Indra Boving and Steven Le"
date: "2026-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor) #clean up datasets nicely
library(here) #helpful for directing R around the computer
```

Read in data: 

Write info here!

Columns are these: 
angle_of_photo_taken: how much the camera was tilted side-to-side. Ignoring if less than 5 degrees.
angle_related_to_straight_line: relative to a vertical line, where < 90 is facing up, and >90 is facing down.
z_axis: 90 = flat leaf is facing camera, 0 = leaf is only viewed from the side


```{r}
#read in the data using read_csv(). Make sure your data is in quotes, and that its in the same folder as your .Rmd file. 
#leafangle_df <- read_csv("scripts/steven/leafangles.csv") %>% 
#  clean_names() 

leafangle_df <- read_csv(here("scripts", "steven", "leafangles.csv")) %>% 
  clean_names() 
  
leafangle_means_df <- leafangle_df %>% 
  group_by(tree_number) %>% #says "cluster my data into groups by tree_id
  mutate(mean_angle = mean(angle_related_to_straight_line),
         sd_angle = sd(angle_related_to_straight_line)) %>% #make a new column called 'mean_angle' and make it the mean value of the angle column, do this per tree (from the group_by)
  select(tree_number, mean_angle, sd_angle) %>% #keep only these two columns, drop the rest
  distinct() #keep only rows that are unique, dont keep duplicates

#look at it: 
leafangle_df
```


Add in the species data: 

```{r}
#read in the csv with the water potential data for each tree: 
tree_wp_df <- read_csv(here("processed-data", "sedgwick_mpas_allyears_20250829.csv")) %>% 
  rename( tree_number = individual_id)

combined_df <- merge(leafangle_means_df, tree_wp_df, by = c("tree_number"))
```

Make some figures: 
```{r}
#use a function called ggplot to make figures
combined_df %>% 
  ggplot(aes(y = mean_angle, 
             x = species,
             fill = species)) +
  geom_point() +
  geom_boxplot(alpha = .3)
```
```{r}
#read in the data using read_csv(). Make sure your data is in quotes, and that its in the same folder as your .Rmd file. 
leafangle_df <- read_csv("scripts/Steven/leafangles.csv") %>% 
  clean_names() 
  
leafangle_means_df <- leafangle_df %>% 
  group_by(tree_number) %>% #says "cluster my data into groups by tree_id
  mutate(mean_angle = mean(angle_related_to_straight_line),
         sd_angle = sd(angle_related_to_straight_line)) %>% #make a new column called 'mean_angle' and make it the mean value of the angle column, do this per tree (from the group_by)
  select(tree_number, mean_angle, sd_angle) %>% #keep only these two columns, drop the rest
  distinct() #keep only rows that are unique, dont keep duplicates

#look at it: 
leafangle_df
```

```{r}
#read in the csv with the water potential data for each tree: 
tree_wp_df <- read_csv(here("processed-data", "sedgwick_mpas_allyears_20250829.csv")) %>% 
  rename( tree_number = individual_id)

combined_df <- merge(leafangle_means_df, tree_wp_df, by = c("tree_number"))
```


#Plots

```{r}
#use a function called ggplot to make figures
combined_df %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = species)) +
  geom_point() +
  geom_boxplot(alpha = .3)
```


```{r}
combined_df %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             fill = mean_angle)) +
  geom_col(width = 1)
```
```{r}
combined_df %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = pd_md)) +
  geom_point() +
  facet_wrap(~date)
```
We need to make a column for week or month!

```{r}
combined_df_time <- combined_df %>% 
  mutate(date_formated = ymd(date),
         month = month(date_formated), 
         year = year(date_formated)) %>% 
  mutate(angle_absolute = case_when(
         mean_angle > 90 ~ 90 - (mean_angle - 90), 
         TRUE ~ as.numeric(mean_angle)))

str(combined_df_time)
```

Middays vs. leaf angle

```{r}
combined_df_time %>% 
  filter(year %in% c(2024, 2025),
         pd_md == "md") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = pd_md, 
             shape = species)) +
  geom_point() +
  facet_wrap(~year*month) +
  geom_smooth(method = "lm")
```

```{r}
combined_df_time %>% 
  filter(year %in% c(2024, 2025),
         pd_md == "md") %>% 
  ggplot(aes(x = angle_absolute, 
             y = water_potential_mean, 
             color = pd_md, 
             shape = species)) +
  geom_point() +
  facet_wrap(~year*month) +
  geom_smooth(method = "lm")
```

Predawns vs. leaf angle: 

```{r}
combined_df_time %>% 
  filter(year %in% c(2024, 2025),
         pd_md == "pd") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = pd_md, 
             shape = species)) +
  geom_point() +
  facet_wrap(~year*month, scales = "free") +
  geom_smooth(method = "lm")
```
```{r}
combined_df_time %>% 
  filter(year %in% c(2024, 2025),
         pd_md == "pd") %>% 
  ggplot(aes(x = angle_absolute, 
             y = water_potential_mean, 
             color = pd_md, 
             shape = species)) +
  geom_point() +
  facet_wrap(~year*month, scales = "free") +
  geom_smooth(method = "lm")
```

#####To do: 
- Make the water potential figure prettier, easier to read   DONE
- Make versions for different months   DONE
- Make the leaf angle and water potentials plots prettier   DONE
- If time, add more measurements, but not necessarily needed   X 
- Summarize the papers youve read (6), 2-3 sentences each and how its relevant for your project   DONE

#####April 2: 

- Double check that sept is actually teh driest for each year
- Make side-by-side graphs of each year/month that are interesting to compare across years
- Run statistical tests on leaf angle ~ predawn (see bwlow)
- Write abstract draft by 2/8, attend workshop if you can! See email from Lee. 

- Think about senior thesis and using LiDar data to get leaf angle, if you want :) 

```{r}
#example stats: 
df <- combined_df_time %>% 
  filter() #filter the month and year you want to compare within, also filter to just predawns

model <- lm(leaf_angle_mean ~ water_potential_mean, data = df)
summary(model) #look for p < 0.05 for significance
```


```{r}
#use a function called ggplot to make figures
combined_df_time %>% 
  filter(month %in% c(9) & year %in% c(2024)) %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = pd_md)) +
  geom_point() +
  geom_boxplot(alpha = .3)
```

```{r}
combined_df_time %>% 
  filter(month %in% c(9) & year %in% c(2024)) %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = pd_md)) +
  geom_point() +
  geom_boxplot(alpha = .3) + 
  labs(y = "Mean Water Potential",
       x = "Species", 
       fill = "Time")
```

```{r}
combined_df_time %>% 
  filter(month %in% c(9) & year %in% c(2024)) %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = pd_md)) +
  geom_point() +
  geom_boxplot(alpha = .3) + 
  labs(y = "Mean Water Potential",
       x = "Species", 
       fill = "Time") + 
  theme(panel.background = element_rect(colour='black'))
```

```{r}
combined_df_time %>% 
  filter(month %in% c(9) & year %in% c(2024)) %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = pd_md)) +
  geom_point() + 
  geom_boxplot(alpha = .3) + 
  scale_fill_manual(
    values = c("pd" = "#4682B4", "md" = "lightcoral"), 
    labels = c("pd" = "Predawn", "md" = "Midday")) +
  labs(y = "Mean Water Potential",
       x = "Species", 
       fill = "Time:") + 
  theme(
panel.background = element_rect(colour = 'black'),     
text = element_text(family = "Times New Roman"), 
    legend.position = "right",                    
    axis.title.y = element_text(size = 25, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 25, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12))
#Water Potential Midday vs. Predawn Figure 1#
```

```{r}
combined_df_time %>% 
  filter(month %in% c(9) & year %in% c(2023)) %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = pd_md)) +
  geom_point() + 
  geom_boxplot(alpha = .3) + 
  scale_fill_manual(
    values = c("pd" = "#4682B4", "md" = "lightcoral"), 
    labels = c("pd" = "Predawn", "md" = "Midday")) +
  labs(y = "Mean Water Potential",
       x = "Species", 
       fill = "Time:") + 
  theme(
panel.background = element_rect(colour = 'black'),     
text = element_text(family = "Times New Roman"), 
    legend.position = "right",                    
    axis.title.y = element_text(size = 25, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 25, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12))
#Water Potential Midday vs. Predawn Figure 2#
```

```{r}
combined_df_time %>% 
  filter(month %in% c(9) & year %in% c(2022)) %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = pd_md)) +
  geom_point() + 
  geom_boxplot(alpha = .3) + 
  scale_fill_manual(
    values = c("pd" = "#4682B4", "md" = "lightcoral"), 
    labels = c("pd" = "Predawn", "md" = "Midday")) +
  labs(y = "Mean Water Potential",
       x = "Species", 
       fill = "Time:") + 
  theme(
panel.background = element_rect(colour = 'black'),     
text = element_text(family = "Times New Roman"), 
    legend.position = "right",                    
    axis.title.y = element_text(size = 25, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 25, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12))
#Water Potential Midday vs. Predawn Figure 3#
```

```{r}
combined_df_time %>% 
filter(year == 2024,
         month == "9",         
       pd_md == "md") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = species, 
             shape = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Quercus agrifolia" = "#FFCC00", 
                                "Quercus douglasii" = "darkgreen")) + 
  labs(y = "Mean Water Potential",
       x = "Mean Angle (Degrees)", 
       title = "Water Potential at Midday",
       subtitle = "September 2024",
       color = "Species:",
       shape = "Species:") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(color = "none") + 
  theme(panel.background = element_rect(colour = 'black'),   
        text = element_text(family = "Times New Roman"),
  axis.title.y = element_text(size = 17, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12),
plot.title = element_text(hjust = 0.5, size = 25, face = "bold", margin = margin(b = 10)),
plot.subtitle = element_text(hjust = 0.5, size = 20, margin = margin(b = 10)))
#Water Potential at Midday Figure 1#
```

```{r}
combined_df_time %>% 
filter(year == 2023,
         month == "9",         
       pd_md == "md") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = species, 
             shape = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Quercus agrifolia" = "#FFCC00", 
                                "Quercus douglasii" = "darkgreen")) + 
  labs(y = "Mean Water Potential",
       x = "Mean Angle (Degrees)", 
       title = "Water Potential at Midday",
       subtitle = "September 2023",
       color = "Species:",
       shape = "Species:") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(color = "none") + 
  theme(panel.background = element_rect(colour = 'black'),   
        text = element_text(family = "Times New Roman"),
  axis.title.y = element_text(size = 17, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12),
plot.title = element_text(hjust = 0.5, size = 25, face = "bold", margin = margin(b = 10)),
plot.subtitle = element_text(hjust = 0.5, size = 20, margin = margin(b = 10)))
#Water Potential at Midday Figure 2#
```

```{r}
combined_df_time %>% 
filter(year == 2022,
         month == "9",         
       pd_md == "md") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = species, 
             shape = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Quercus agrifolia" = "#FFCC00", 
                                "Quercus douglasii" = "darkgreen")) + 
  labs(y = "Mean Water Potential",
       x = "Mean Angle (Degrees)", 
       title = "Water Potential at Midday",
       subtitle = "September 2022",
       color = "Species:",
       shape = "Species:") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(color = "none") + 
  theme(panel.background = element_rect(colour = 'black'),   
        text = element_text(family = "Times New Roman"),
  axis.title.y = element_text(size = 17, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12),
plot.title = element_text(hjust = 0.5, size = 25, face = "bold", margin = margin(b = 10)),
plot.subtitle = element_text(hjust = 0.5, size = 20, margin = margin(b = 10)))
#Water Potential at Midday Figure 3#
```

```{r}
combined_df_time %>% 
filter(year == 2024,
         month == "9",         
       pd_md == "pd") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = species, 
             shape = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Quercus agrifolia" = "#FFCC00", 
                                "Quercus douglasii" = "darkgreen")) + 
  labs(y = "Mean Water Potential",
       x = "Mean Angle (Degrees)", 
       title = "Water Potential at Predawn",
       subtitle = "September 2024",
       color = "Species:",
       shape = "Species:") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(color = "none") + 
  theme(panel.background = element_rect(colour = 'black'),   
        text = element_text(family = "Times New Roman"),
  axis.title.y = element_text(size = 17, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12),
plot.title = element_text(hjust = 0.5, size = 25, face = "bold", margin = margin(b = 10)),
plot.subtitle = element_text(hjust = 0.5, size = 20, margin = margin(b = 10)))
#Water Potential at Predawn Figure 1#
```

```{r}
combined_df_time %>% 
filter(year == 2023,
         month == "9",         
       pd_md == "pd") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = species, 
             shape = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Quercus agrifolia" = "#FFCC00", 
                                "Quercus douglasii" = "darkgreen")) + 
  labs(y = "Mean Water Potential",
       x = "Mean Angle (Degrees)", 
       title = "Water Potential at Predawn",
       subtitle = "September 2023",
       color = "Species:",
       shape = "Species:") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(color = "none") + 
  theme(panel.background = element_rect(colour = 'black'),   
        text = element_text(family = "Times New Roman"),
  axis.title.y = element_text(size = 17, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12),
plot.title = element_text(hjust = 0.5, size = 25, face = "bold", margin = margin(b = 10)),
plot.subtitle = element_text(hjust = 0.5, size = 20, margin = margin(b = 10)))
#Water Potential at Predawn Figure 2#
```

```{r}
combined_df_time %>% 
filter(year == 2022,
         month == "9",         
       pd_md == "pd") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = species, 
             shape = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Quercus agrifolia" = "#FFCC00", 
                                "Quercus douglasii" = "darkgreen")) + 
  labs(y = "Mean Water Potential",
       x = "Mean Angle (Degrees)", 
       title = "Water Potential at Predawn",
       subtitle = "September 2022",
       color = "Species:",
       shape = "Species:") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(color = "none") + 
  theme(panel.background = element_rect(colour = 'black'),   
        text = element_text(family = "Times New Roman"),
  axis.title.y = element_text(size = 17, margin = margin(r = 10)), 
    axis.title.x = element_text(size = 17, margin = margin(t = 10)),
    axis.text = element_text(size = 12),
legend.title = element_text(size = 15),
    legend.text = element_text(size = 12),
plot.title = element_text(hjust = 0.5, size = 25, face = "bold", margin = margin(b = 10)),
plot.subtitle = element_text(hjust = 0.5, size = 20, margin = margin(b = 10)))
#Water Potential at Predawn Figure 3#
```

