---
title: "Leaf Angle"
author: "Indra Boving and Steven Le"
date: "2026-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor) #clean up datasets nicely
library(here) #helpful for directing R around the computer
```

Read in data: 

Write info here!

Columns are these: 
angle_of_photo_taken: how much the camera was tilted side-to-side. Ignoring if less than 5 degrees.
angle_related_to_straight_line: relative to a vertical line, where < 90 is facing up, and >90 is facing down.
z_axis: 90 = flat leaf is facing camera, 0 = leaf is only viewed from the side


```{r}
#read in the data using read_csv(). Make sure your data is in quotes, and that its in the same folder as your .Rmd file. 
leafangle_df <- read_csv("leafangles.csv") %>% 
  clean_names() 
  
leafangle_means_df <- leafangle_df %>% 
  group_by(tree_number) %>% #says "cluster my data into groups by tree_id
  mutate(mean_angle = mean(angle_related_to_straight_line),
         sd_angle = sd(angle_related_to_straight_line)) %>% #make a new column called 'mean_angle' and make it the mean value of the angle column, do this per tree (from the group_by)
  select(tree_number, mean_angle, sd_angle) %>% #keep only these two columns, drop the rest
  distinct() #keep only rows that are unique, dont keep duplicates

#look at it: 
leafangle_df
```


Add in the species data: 

```{r}
#read in the csv with the water potential data for each tree: 
tree_wp_df <- read_csv(here("processed-data", "sedgwick_mpas_allyears_20250829.csv")) %>% 
  rename( tree_number = individual_id)

combined_df <- merge(leafangle_means_df, tree_wp_df, by = c("tree_number"))
```

Make some figures: 
```{r}
#use a function called ggplot to make figures
combined_df %>% 
  ggplot(aes(y = mean_angle, 
             x = species,
             fill = species)) +
  geom_point() +
  geom_boxplot(alpha = .3)
```
```{r}
#read in the data using read_csv(). Make sure your data is in quotes, and that its in the same folder as your .Rmd file. 
leafangle_df <- read_csv("leafangles.csv") %>% 
  clean_names() 
  
leafangle_means_df <- leafangle_df %>% 
  group_by(tree_number) %>% #says "cluster my data into groups by tree_id
  mutate(mean_angle = mean(angle_related_to_straight_line),
         sd_angle = sd(angle_related_to_straight_line)) %>% #make a new column called 'mean_angle' and make it the mean value of the angle column, do this per tree (from the group_by)
  select(tree_number, mean_angle, sd_angle) %>% #keep only these two columns, drop the rest
  distinct() #keep only rows that are unique, dont keep duplicates

#look at it: 
leafangle_df
```

```{r}
#read in the csv with the water potential data for each tree: 
tree_wp_df <- read_csv(here("processed-data", "sedgwick_mpas_allyears_20250829.csv")) %>% 
  rename( tree_number = individual_id)

combined_df <- merge(leafangle_means_df, tree_wp_df, by = c("tree_number"))
```


#Plots

```{r}
#use a function called ggplot to make figures
combined_df %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = species)) +
  geom_point() +
  geom_boxplot(alpha = .3)
```


```{r}
combined_df %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             fill = mean_angle)) +
  geom_col(width = 1)
```
```{r}
combined_df %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = pd_md)) +
  geom_point() +
  facet_wrap(~date)
```
We need to make a column for week or month!

```{r}
combined_df_time <- combined_df %>% 
  mutate(date_formated = ymd(date),
         month = month(date_formated), 
         year = year(date_formated)) %>% 
  mutate(angle_absolute = case_when(
         mean_angle > 90 ~ 90 - (mean_angle - 90), 
         TRUE ~ as.numeric(mean_angle)))

str(combined_df_time)
```

Middays vs. leaf angle

```{r}
combined_df_time %>% 
  filter(year %in% c(2024, 2025),
         pd_md == "md") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = pd_md, 
             shape = species)) +
  geom_point() +
  facet_wrap(~year*month) +
  geom_smooth(method = "lm")
```

```{r}
combined_df_time %>% 
  filter(year %in% c(2024, 2025),
         pd_md == "md") %>% 
  ggplot(aes(x = angle_absolute, 
             y = water_potential_mean, 
             color = pd_md, 
             shape = species)) +
  geom_point() +
  facet_wrap(~year*month) +
  geom_smooth(method = "lm")
```

Predawns vs. leaf angle: 

```{r}
combined_df_time %>% 
  filter(year %in% c(2024, 2025),
         pd_md == "pd") %>% 
  ggplot(aes(x = mean_angle, 
             y = water_potential_mean, 
             color = pd_md, 
             shape = species)) +
  geom_point() +
  facet_wrap(~year*month, scales = "free") +
  geom_smooth(method = "lm")
```
```{r}
combined_df_time %>% 
  filter(year %in% c(2024, 2025),
         pd_md == "pd") %>% 
  ggplot(aes(x = angle_absolute, 
             y = water_potential_mean, 
             color = pd_md, 
             shape = species)) +
  geom_point() +
  facet_wrap(~year*month, scales = "free") +
  geom_smooth(method = "lm")
```

#####To do: 
- Make the water potential figure prettier, easier to read
- Make versions for different months
- Make the leaf angle and water potentials plots prettier
- If time, add more measurements, but not necessarily needed 
- Summarize the papers youve read (6), 2-3 sentences each and how its relevant for your project. 

```{r}
#use a function called ggplot to make figures
combined_df_time %>% 
  filter(month %in% c(9) & year %in% c(2024)) %>% 
  ggplot(aes(y = water_potential_mean, 
             x = species,
             fill = pd_md)) +
  geom_point() +
  geom_boxplot(alpha = .3)
```